Data-driven measure of convergence for word embeddings

### 2017-07-07 draft2.tex
Changes based on brainstorming session

\section{Intoduction}
\label{sec:introduction}
Since their introduction \cite{Mikolov_CS2013}, word embeddings have become increasingly popular and have been used for a wide range of applications \cite{Mikolov_CoRR2013, Kim_CoRR2014, Levy_CNLL2014, Cordeiro_ACL2016, Smith_ICLR2017}. Word embeddings are often trained using large data sets, such as the Google News corpus [https://code.google.com/archive/p/word2vec/], incorporating 100 billion words. The generally accepted approach when training word embeddings is "the more data, the better", and there does not appear to be any consensus on the minimum amount of data required to train a word embedding model which yields good results. A widely accepted method for evaluating word embeddings is by testing their performance in analogy tests: \textit{man : woman :: king : \textbf{queen}}. Data sets for testing analogies are widely available (see 
\url{https://www.aclweb.org/aclwiki/index.php?title=Google_analogy_test_set_(State_of_the_art)}). However there several known issues with such tests \cite{Speer_Blog2016}:

\begin{itemize}
 \item \textbf{Language specific} - analogies are usually language specific -- training a word embedding model on a language different from that of the analogies will obviously lead to failing the test;
 \item \textbf{Domain specific} - analogies usually fall in a particular knowledge category. For instance the following analogy: \textit{Athens : Greece :: Baghdad : Iraq}, is specific to Geographical facts; a semantic model trained on a data set which does not contain such knowledge (which never mentions these locations) is bound to fail this test. At the same time, a word embedding may have already learnt a domain specific analogy, such as \textit{electron : electricity :: photon : light}, but if this analogy is not reflected on the analogy test set, there is no way to know the semantic model already incorporates this knowledge;
 \item \textbf{Constrained by vocabulary} - because word embeddings typically have a set vocabulary; analogies which use out-of-vocabulary words, will always fail. Domain and language specific issues can be considered specific instances of constrained vocabularies.
\end{itemize}

These issues can be addressed, for instance, by translating the analogy test set, discarding analogies which are irrelevant for the training data being used, constructing analogies suitable for a domain, etc. However all of these approaches are labour intensive and must be hand crafted for each training data set. This work is motivated by the desire to find a data-driven method to validate word embeddings. The data-driven method introduced here does not rely on the vocabulary and, unlike the analogy test, does not require translation or creation of new test sets in order to function. This makes our approach effectively language independent. This may also be particularly useful in cases where the available amount of data for a particular domain is relatively small.

@@

@@

\begin{algorithm}[H]
% \KwData{this text}
% \KwResult{how to write algorithm with \LaTeX2e }
% initialization\;
% \While{not at end of this document}{
%  read current\;
%  \eIf{understand}{
%   go to next section\;
%   current section becomes this one\;
%   }{
%   go back to the beginning of current section\;
%  }
% }
  \ForEach{ dimension $i$ in the embedded space }{
    $V1_i$ $\leftarrow$ a unit vector along the i-th dimension of the embedded space $S_A$ \;
    $W_i$ $\leftarrow$ $V1_i$ projected back into vocabulary space \footnote{To convert $V1_i$ back into vocabulary space, we take the product of the embedded vector and the inverse of embedding matrix.} \footnote{If the vocabularies are different we need to convert $W_i$ to the other vocabulary} \;
    $V2_i$ $\leftarrow$ $W_i$ projected in embedded space $S_B$ \;
    $V2_i$ $\leftarrow$ the i-th dimension of the transformation matrix $T$ \;
  }
 \caption{Find $T$ for $S_A$ to $S_B$}
\end{algorithm}

@@

$$T_i = S_B S_A^{-1} V1_i $$

where $V1_i$ is the unit vector along dimension i in the semantic space defined by $S_A$.

@@

\section{Experiments}
\label{sec:experiments}
In this section we describe the experimental process we have followed to validate our approach. For this experiment we used the Times news paper archives (\textbf{I need to ask Jose how can we reference the corpus}) between 1900 and 1980. We trained a semantic model, using gensim word2vec implementation [ref]. We set the dimensions on the semantic space to 300 ($N = 300$).

@@

For comparison, we also use the Google "questions-words" analogy data set [ref]. It should be mentioned that some of the analogy tests failed for every one of our trained models because some words in those tests were not present in our corpus (more specifically, they were not present with enough frequency to be part of the vocabulary in our models). For example, even after training with 10 million sentences, our model fails for the analogy: \textit{London : England :: Montevideo : Uruguay}. This is however not surprising as \textit{Montevideo} appears only a 87 times in the first million sentences (for reference, London appears in 21,000 sentences). Notice that the proposed measure calculates the difference between two embeddings, while the question words evaluates the individual accuracy of a given model.

@@

Figure \ref{fig:1} shows the convergence score (blue) as the number of sentences increases, and the Google accuracy (green). After $10^6$ sentences have been used for training the model, there is a rapid increase in the convergence score. This indicates that the vectors defining the semantic space are becoming more stable. Training the model with additional sentences will produce little change in the model.

@@

\begin{figure}
 \includegraphics[scale=1]{figure1.png} 
 \caption{xxx}
 \label{fig:1}
\end{figure}

@@

Figure \ref{fig:2} shows the values on the diagonal of diagonal convergence scores as the number of sentences increases. Each line represents the convergence score for one particular dimension of the semantic space. What we see here is that around $2 * 10^4$ sentences there is wide variety of values for the convergence scores -- some dimensions of the semantic space are relatively stable (high convergence scores) while others are undergoing significant change (low convergence scores). As more sentences are added to the model, the changes in semantic model become more widespread, until finally the model becomes globally stable.

\begin{figure}
 \includegraphics[scale=1]{figure2.png} 
 \caption{xxx}
 \label{fig:2}
\end{figure}

@@

\begin{figure}
 \includegraphics[scale=1]{figure3.png} 
 \caption{xxx}
 \label{fig:3}
\end{figure}

@@

The convergence scores can be visualized as a single value over time, as shown on Figure 1, or as a collection of values as shown on Figure 2. The first form of visualization is more straight forward, while the second provides additional insight into the structure of the semantic space as a whole.

@@

\section{Notes}
\begin{itemize}
\item Add a clarification (definition) of what it means for two spaces to be equivalent:
  \begin{itemize}
  \item Relations in one space are preserved in the other
  \item It is basically just a rotation
  \end{itemize}

 \item "Known issues:" restructure (maybe into one thing)

 \item  Pseudocode 
 \begin{itemize}
  \item Add picture (how would it look like)
  \item Add intuitive / concrete example (using a word)
 \end{itemize}

 \item Separate repo for the code

 \item Move "We use D(A,B)..." to Method section (together with clarification of definition above)
 \item Add: we used CBOW (or skip gram)

 \item Figure 2 
 \begin{itemize}
  \item density plot
  \item xaxis should be No. sentences
 \end{itemize}

 \item Figure 3
  \begin{itemize}
  \item add axis No. sentences
  \item add discussion "why is google score low"
 \end{itemize}

 \item Add in introduction why we do this more strongly

 \item Venues:
 \begin{itemize}
  \item LREC
  \item IEEE
  \item NIPS
  \item PR ?
  \item http://cs.rochester.edu/~omidb/nlpcalendar/
 \end{itemize}

\end{itemize}

>>>>

\section{Intoduction}
\label{sec:introduction}
Since their introduction \cite{Mikolov_CS2013}, word embeddings have become increasingly popular and have been used for a wide range of applications \cite{Mikolov_CoRR2013, Kim_CoRR2014, Levy_CNLL2014, Cordeiro_ACL2016, Smith_ICLR2017}. Word embeddings are often trained using large data sets, such as the Google News corpus [https://code.google.com/archive/p/word2vec/], incorporating 100 billion words. The generally accepted approach when training word embeddings is the more data, the better, and there does not appear to be any consensus on the minimum amount of data required to train a word embedding model which yields good results. A widely accepted method for evaluating word embeddings is by testing their performance in analogy tests: \textit{man : woman :: king : \textbf{queen}}. Data sets for testing analogies are widely available (see \url{https://www.aclweb.org/aclwiki/index.php?title=Google_analogy_test_set_(State_of_the_art)}). However there are several known issues with such tests \cite{Speer_Blog2016}:

\begin{itemize}
 \item \textbf{Language specific} - analogies are usually language specific -- training a word embedding model on a language different from that of the analogies will obviously lead to failing the test;
 \item \textbf{Domain specific} - analogies usually fall in a particular knowledge category. For instance the following analogy: \textit{Athens : Greece :: Baghdad : Iraq}, is specific to Geographical facts; a semantic model trained on a data set which does not contain such knowledge (which never mentions these locations) is bound to fail this test. At the same time, a word embedding may have already learnt a domain specific analogy, such as \textit{electron : electricity :: photon : light}, but if this analogy is not reflected on the analogy test set, there is no way to know the semantic model already incorporates this knowledge;
 \item \textbf{Constrained by vocabulary} - because word embeddings typically have a set vocabulary; analogies which use out-of-vocabulary words, will always fail. Domain and language specific issues can be considered specific instances of constrained vocabulary issues.
\end{itemize}

These issues can be addressed, for instance, by translating the analogy test sets, discarding analogies which are irrelevant for the data being used, constructing analogies suitable for a domain, etc. However all of these approaches are labour intensive and must be hand crafted for each training data set. This work is motivated by the desire to find a data-driven method to validate word embeddings. The data-driven method introduced here does not rely on the vocabulary and thus does not require translation or creation of new test sets in order to function. This makes our approach effectively language independent.

@@

This method may also be particularly useful in cases where the available data needs to be divided to allow for diachronic data analysis. In such cases there is a clear trade-off between the volume of data used to train the model and the temporal resolution of the analysis: data slices spanning longer periods of time contain more data and produce better models; on the other hand, data slices spanning shorter periods of time will allow a more detailed temporal resolution. The data-driven method allows us to find the right balance for such scenarios.

@@

\begin{algorithm}[H]
  \ForEach{ dimension $i$ in the embedded space }{
    $\widehat{i_A}$ $\leftarrow$ unit vector along the i-th dimension of the embedded space $S_A$ \;
    $W_i$ $\leftarrow$ $\widehat{i_A}$ projected back into vocabulary space \footnote{To convert $\widehat{i_A}$ back into vocabulary space, we take the product of the embedded vector and the inverse of embedding matrix.} \footnote{If the vocabularies are different we need to convert $W_i$ to the other vocabulary} \;
    $T_i$ $\leftarrow$ $W_i$ projected in embedded space $S_B$ \;
    i-th dimension of $T$ $\leftarrow$ $T_i$ \;
  }
 \caption{Find $T$ to map from $S_A$ to $S_B$}
\end{algorithm}

@@

$$T_i = S_B S_A^{-1} \widehat{i_A} $$

where $\widehat{i_A}$ is the unit vector along dimension $i$ in the semantic space defined by $S_A$.

Figure \ref{fig:transform} depicts how the transformation matrix is built.

\begin{figure}
 \includegraphics[width=\textwidth]{transformMatrix.pdf} 
 \caption{The $i$-th dimension transformation matrix $T$ ($T_i$), represents vector $\widehat{i_A}$ transformed from space defined by $S_A$ to from space defined by $S_B$.}
 \label{fig:transform}
\end{figure}

@@

\section{Experiments}
\label{sec:experiments}
In this section we describe the experimental process we have followed to validate our approach. For this experiment we used the Times news paper archives (\textbf{I need to ask Jose how can we reference the corpus}) between 1900 and 1980. We trained a semantic model, using gensim word2vec implementation, using CBOW [ref]. We set the dimensions on the semantic space to 300 ($N = 300$).

@@

We use $D(A,B)$ to compare snapshots. For comparison, we also use the Google "questions-words" analogy data set [ref]. It should be mentioned that some of the analogy tests failed for every one of our trained models because some words in those tests were not present in our corpus (more specifically, they were not present with enough frequency to be part of the vocabulary in our models). For example, even after training with 10 million sentences, our model fails for the analogy: \textit{London : England :: Montevideo : Uruguay}. This is however not surprising as \textit{Montevideo} appears only a 87 times in the first million sentences (for reference, London appears in 21,000 sentences). Notice that the proposed measure calculates the difference between two embeddings, while the question words evaluates the individual accuracy of a given model.

@@

Figure \ref{fig:scores} shows the convergence score (blue) as the number of sentences increases, and the Google accuracy (green). After $10^6$ sentences have been used for training the model, there is a rapid increase in the convergence score. This indicates that the vectors defining the semantic space are becoming more stable. Training the model with additional sentences will produce little change in the model.

@@

\begin{figure}
 \includegraphics[width=\textwidth]{scoresVsNoSentences.pdf} 
 \caption{xxx}
 \label{fig:scores}
\end{figure}

@@

Figure \ref{fig:spreadDiagonal} shows the values on the diagonal of diagonal convergence scores as the number of sentences increases. Each line represents the convergence score for one particular dimension of the semantic space. What we see here is that around $2 * 10^4$ sentences there is wide variety of values for the convergence scores -- some dimensions of the semantic space are relatively stable (high convergence scores) while others are undergoing significant change (low convergence scores). As more sentences are added to the model, the changes in semantic model become more widespread, until finally the model becomes globally stable. \textbf{TODO -- Discuss: Why is Google Accuracy score so low ?}

\begin{figure}
 \includegraphics[width=\textwidth]{spreadDiagonalValues.pdf} 
 \caption{xxx}
 \label{fig:spreadDiagonal}
\end{figure}

@@

\begin{figure}
 \includegraphics[width=\textwidth]{diachronicSlices.pdf} 
 \caption{xxx}
 \label{fig:diachronic}
\end{figure}

@@

The convergence scores can be visualized as a single value over time, as shown on Figure \ref{fig:scores}, or as a collection of values as shown on Figure \ref{fig:spreadDiagonal}. The first form of visualization is more straight forward, while the second provides additional insight into the structure of the semantic space as a whole.

@@

%\section{Notes}
%\begin{itemize}
%\item Add a clarification (definition) of what it means for two spaces to be equivalent:
%  \begin{itemize}
%  \item Relations in one space are preserved in the other
%  \item It is basically just a rotation
%  \end{itemize}
%   
% \item "Known issues:" restructure (maybe into one thing)
%
% \item  Pseudocode 
% \begin{itemize}
%  \item Add picture (how would it look like)
%  \item Add intuitive / concrete example (using a word)
% \end{itemize}
%
% \item Separate repo for the code
%
% \item Move "We use D(A,B)..." to Method section (together with clarification of definition above)
% \item Add: we used CBOW (or skip gram)
% 
% \item Figure 2 
% \begin{itemize}
%  \item density plot
%  \item xaxis should be No. sentences
% \end{itemize}
%
% \item Figure 3
%  \begin{itemize}
%  \item add axis No. sentences
%  \item add discussion "why is google score low"
% \end{itemize}
%
% \item Add in introduction why we do this more strongly
%
% \item Venues:
% \begin{itemize}
%  \item LREC
%  \item IEEE
%  \item NIPS
%  \item PR ?
%  \item http://cs.rochester.edu/~omidb/nlpcalendar/
% \end{itemize}
%
%\end{itemize}
